---
title: "**Analysis of Online Chess Games**"
author: "**_Manami Das_** \n\t\n _MDS202423_ "
output: 
  pdf_document: 
    toc_depth: 0
    fig_width: 4
    fig_height: 2
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

This project presents an **Analysis of Online Chess played on Lichess** , focusing on various matches with details like the winner, rating of players, number of moves etc. The study explores the effects of rating and opening moves on the result of match and the comparison of results according to the piece colour.

---

## Introduction

Chess is a board game for two players. It is an abstract strategy game that involves no hidden information or elements of chance. It is played on a chessboard with 64 squares arranged in an *8×8 grid*. The players, referred to as "**White**" and "**Black**", each control sixteen pieces: one king, one queen, two rooks, two bishops, two knights, and eight pawns. White moves first, followed by Black. The game is typically won by checkmating the opponent's king, i.e. threatening it with inescapable capture. There are several ways a game can end in a draw.

**Lichess** is a free and open-source Internet chess server.**Lichess** was founded in *2010* by French programmer _Thibault Duplessis_.Users of the site can play online chess anonymously and optionally register an account to play rated games .

---

## Data Description

The **Online Chess Games** gives data for 20,058 chess games played on *Lichess*, including moves, victor, rating, opening details and more. Below is the list of column names:
\begin{table}[ht]
\centering
\begin{tabular}{lcl}
  \hline
 & Column Name & Description  \\ 
  \hline
  1 & game\_id & Unique id for each match  \\ 
  2 & rated & To show whether a match is rated or not \\ 
  3 & turns & To give total number of turns in a match played \\ 
  4 & victory\_status & To show the result of match is due to Mate, Resign, Draw or Time Out \\ 
  5 & winner & To give winner piece colour  \\ 
  6 & white\_id & To show the id of player with white piece \\ 
  7 & white\_rating & To give the rating of player with white piece\\ 
  8 & black\_id & To show the id of player with black piece \\ 
  9 & black\_rating & To give the rating of player with black piece\\ 
  10 & moves & To give the list of complete set of moves played in a match \\ 
   \hline
\end{tabular}
\caption{Columns in Dataset} 
\end{table}

## Exploratory Data Analysis
```{r libraries, include=FALSE}
##Required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
```

```{r import, include=FALSE}
#Importing File and naming the columns 
chess<-read.csv("C:/Users/pcm/Desktop/cmi/Project/Visu/chess_game.csv")
id=chess$game_id
rating=chess$rated
turns=chess$turns
status=chess$victory_status
winner=chess$winner
white_id=chess$white_id
black_id=chess$black_id
white_rating=chess$white_rating
black_rating=chess$black_rating
moves=chess$moves
opening_code=chess$opening_code
opening_moves=chess$opening_moves
```
### Percentage of Rated Matches:
```{r no. of matches , include=FALSE}
#total match and rated matches
total_matches=nrow(chess)
rating_tally=table(rating)
print(total_matches)
```

```{r Rated vs Unrated ,echo=FALSE, fig.width=2.25,fig.height=1.5,fig.align="center",fig.cap="Rated vs Unrated Matches"}
df <- data.frame(Category = c("Rated", "Unrated"), value = c(16155, 3903))
ggplot(df, aes(x = "", y = value, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +geom_text(aes(label = paste0(format(round((value*100)/total_matches, 2), nsmall = 2), "%")), position = position_stack(vjust=0.6),size=3) +
  coord_polar("y", start = 0) +
  theme_void()
```
There are 80.54\% rated matches and 19.46\% unrated matches amongst the data. According to the site, the ratings of player is affected by rated matches, the unrated matches, also called the casual matches, do not have any effect on the ratings. A new user has 1500 as their initial rating which increases or decreases  according to their performance in the rated matches.

### Players who played most number of matches:
```{r Top Players,include=FALSE}
chess_players=c(white_id,black_id)
chess_player.tally=table(chess_players)
chess_player.tally_sorted=sort(chess_player.tally,decreasing=TRUE)
no._of_unique_players=length(chess_player.tally)
no_of_chess_players=length(chess_players)
print(no._of_unique_players)
print(no_of_chess_players)
max_plays=chess_player.tally_sorted[1:5]
print(max_plays)
```
```{r Stats of top players, include=FALSE}
dataframe<-data_frame(white_id,black_id,winner)
taranga<-dataframe[((dataframe$white_id=="taranga")&(dataframe$winner=="White"))|((dataframe$black_id=="taranga")&(dataframe$winner=="Black")),]
no._taranga=nrow(taranga)
a_p_t_e_m_u_u<-dataframe[((dataframe$white_id=="a_p_t_e_m_u_u")&(dataframe$winner=="White"))|((dataframe$black_id=="a_p_t_e_m_u_u")&(dataframe$winner=="Black")),]
no._a_p_t_e_m_u_u=nrow(a_p_t_e_m_u_u)
vladimir_kramnik_1<-dataframe[((dataframe$white_id=="vladimir-kramnik-1")&(dataframe$winner=="White"))|((dataframe$black_id=="vladimir-kramnik-1")&(dataframe$winner=="Black")),]
no.vladimir_kramnik_1=nrow(vladimir_kramnik_1)
king5891<-dataframe[((dataframe$white_id=="king5891")&(dataframe$winner=="White"))|((dataframe$black_id=="king5891")&(dataframe$winner=="Black")),]
no.king5891=nrow(king5891)
saviter<-dataframe[((dataframe$white_id=="saviter")&(dataframe$winner=="White"))|((dataframe$black_id=="saviter")&(dataframe$winner=="Black")),]
no.saviter=nrow(saviter)

print(no._taranga/154)
print(no._a_p_t_e_m_u_u/96)
print(no.vladimir_kramnik_1/96)
print(no.king5891/78)
print(no.saviter/76)
```

There are 15635 unique players participating in the data given. The top 5 players who played the most are:

|Player ID|No. matches played|No. of Matches Won|Percentage of Winning
|---------|------------------|------------------|---------
|taranga  |154               |72                |46.75\%
|vladimir\-kramnik\-1|96     |50                |52.08\%
|a\_p\_t\_m\_u\_u|96         |46                |47.92\%
|king5891 |78                |42                |53.85\%
|saviter  |76                |32                |42.11\%

### Most common Opening Moves:
```{r common moves, include=FALSE}
first_moves=rep(NA,total_matches)
moves_list=NA
for (i in 1:total_matches){
  moves_list=unlist(strsplit(moves[i], "\\s+"))
  first_moves[i]=moves_list[1]
  }
unique_first_moves=unique(first_moves)
type=table(first_moves)

sorted=sort(type,decreasing=TRUE)
##5 topmost common moves
for (i in 1:5){
print(sorted[i])}
rest=total_matches-(12598+4522+725+716+416)
print(rest)
```

```{r Most Common Moves,echo=FALSE,fig.width=6,fig.height=2.4,fig.align="center",fig.cap="Most Common Opening Moves"}
Opening_Move = c("1.King's Pawn(e4)","2.Queen's Pawn(d4)", "3.Zukertort Opening(Nf3)","4.English Opening(c4)","5.Van 't Kruijs Opening(e3)",".Other")
value_moves = c(12598,4522,725,716,416,1081)
df2 <- data.frame(Opening_Move , value_moves)
ggplot(df2, aes(x = "", y = value_moves,fill=Opening_Move)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +geom_text(aes(label = paste0(format(round((value_moves*100)/total_matches, 0), nsmall = 0), "%")), position = position_stack(vjust=.4),size=2.5) +
  labs(x =Opening_Move , y ="No. of matches")+
  theme_void()+ theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank()) +
  scale_fill_brewer(palette="Set2")
```
The figure 2 shows that *King's Pawn* is the most commonly used opening move.

### Number of Turns in a Match:
```{r Number of Turns, include=FALSE}
Title=c("Maximum","Minimum","Mean","Median")
V=cbind(max(turns),min(turns),mean(turns),median(turns))
Turns_tally=rbind(Title,V)
```
The summery of number of turns in a match is given below:

|Maximum|Minimum|Mean     |Median
|-------|-------|---------|-----
|349    |1      |60.466   |55

### Percentage of winners with respect to piece colours:
```{r Table of winners,include=FALSE}
winnerpercolour=table(winner)
```

```{r Winner per colour,echo=FALSE, fig.width=5,fig.height=2,fig.align="center",fig.cap="Winners per coloured piece"}
Winner = c("White","Black","Draw")
value_wonby = c(10001,9107,950)
df3 <- data.frame(Winner , value_wonby)
ggplot(df3, aes(x = "", y = value_wonby,fill=Winner)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +geom_text(aes(label = paste0(format(round((value_wonby*100)/total_matches,0), nsmall = 0), "%")), position = position_stack(vjust=0.55),size=3) +
  labs(x ="Winner" , y ="No. of matches")+
  theme_void()+ theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank())+
 scale_fill_brewer(palette="Accent")

```
The figure 3 shows the percentage of different coloured pieces winning matches. It is seen that by less margin the player starting first wins more.

### Different Victory status percentage:
```{r Table of victory status,include=FALSE}
victorystatus=table(status)
```

```{r Victory Status,echo=FALSE, fig.width=5,fig.height=2,fig.align="center",fig.cap="Victory Status"}
Victory_Status = c("Draw","Mate","Out of Time","Resign")
value_status = c(906,6325,1680,11147)
df4 <- data.frame(Victory_Status, value_status)
ggplot(df4, aes(x = "", y = value_status,fill=Victory_Status)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +geom_text(aes(label = paste0(format(round((value_status*100)/total_matches,0), nsmall = 0), "%")), position = position_stack(vjust=0.45),size=3) +
  labs(x ="Victory Status" , y ="No. of matches")+
  theme_void()+ theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank())+
 scale_fill_hue(c = 80,h=c(0,360),l=85)
```
The figure 4 shows that 56\% matches got over due to resignation by a player (this gives the reason of only one move in a match) and there are 8\% of matches which got over due to running out of time (this gives the occurrence of moves as many as 300+ moves).

### First Moves by Winners:
```{r First Moves Tally, include=FALSE}
first_move_of_winner=table(winner,first_moves)
whitewin=sort(first_move_of_winner[3,],decreasing = T)
print(whitewin[1:5])
blackwin=sort(first_move_of_winner[1,],decreasing = T)
print(blackwin[1:5])
drawin=sort(first_move_of_winner[2,],decreasing = T)
print(drawin[1:5])
rest_white=sum(whitewin[6:20])
rest_black=sum(blackwin[6:20])
rest_draw=sum(drawin[6:20])
```

```{r Percentages of Moves, include=FALSE}
whiteW=10001
blackW=9107
drawW=950
div=c(whiteW,blackW,drawW)
F.King.s_Pawn_e4=c(6371,5651,576)*100/div #percentage
E.Queen.s_Pawn_d4=c(2258,2043,221)*100/div #percentage
D.English_Opening_c4=c(383,305,28)*100/div #percentage
C.Zukertort_Opening_Nf3=c(373,292,60)*100/div #percentage
B.Van.t_Kruijs_Opening_e3=c(142,255,19)*100/div #percentage
A.Rest_moves=c(rest_white,rest_black,rest_draw)*100/div #percentage
```
```{r First Moves,echo=FALSE, fig.width=5,fig.height=3.5,fig.align="center",fig.cap="First Moves by Winners"}
Winner_of_match=c("White","Black","Draw")
data_for_opening=data.frame(A.Rest_moves,B.Van.t_Kruijs_Opening_e3,C.Zukertort_Opening_Nf3,D.English_Opening_c4,E.Queen.s_Pawn_d4,F.King.s_Pawn_e4,Winner_of_match)
data_for_opening %>% 
  pivot_longer(
    cols = -Winner_of_match,
    names_to = "Moves",
    values_to = "Percentage", 
    values_transform = list(Percentage=as.integer)
  ) %>% 
 ggplot(aes(x =Winner_of_match, y=Percentage,fill=Moves))+
  geom_col(position = position_fill()) +
  scale_y_continuous(labels = scales::percent) +
  geom_text(aes(label = paste0(format(round((Percentage),0), nsmall = 0), "%")),,
            position = position_fill(vjust = .5),size=2)+
  scale_fill_manual(
    values = c( "lightblue","plum3","lightgreen", "wheat3","peachpuff", "lightpink")
  )
```
The figure 5 shows that irrespective of colour of the winner, the move King's Pawn is preferred as an opening move  followed by Queen's Pawn and English Opening.

### Winner as per ratings:

```{r Winner for rating Tally, include=FALSE}
maxrating=max(white_rating,black_rating)
summary(white_rating,black_rating)
data_rating=data_frame(white_rating,black_rating,winner)
data_rating$more=ifelse(data_rating$white_rating > data_rating$black_rating, 'White', 
                      ifelse(data_rating$white_rating < data_rating$black_rating, 'Black',"Draw"))
data_rating_result= data_rating[data_rating$winner == data_rating$more, ]
moreiswinner=nrow(data_rating_result)##More rated is winner
perc_more_is_winner=moreiswinner/total_matches*100
data_rating_result_opp=data_rating[(data_rating$winner != data_rating$more)&(data_rating$winner!="Draw"),]
lessiswinner=nrow(data_rating_result_opp)
perc_less_is_winner=lessiswinner/total_matches*100
perc_of_draw=100-perc_less_is_winner-perc_more_is_winner
data_for_white_more=data_rating_result[data_rating_result$winner=="White",]
white_for_more=nrow(data_for_white_more)/total_matches*100
data_for_black_more=data_rating_result[data_rating_result$winner=="Black",]
black_for_more=nrow(data_for_black_more)/total_matches*100
data_for_white_less=data_rating_result_opp[data_rating_result_opp$winner=="White",]
white_for_less=nrow(data_for_white_less)/total_matches*100
data_for_black_less=data_rating_result_opp[data_rating_result_opp$winner=="Black",]
black_for_less=nrow(data_for_black_less)/total_matches*100
```

```{r Winner Rating,echo=FALSE, fig.width=5.65,fig.height=2.5,fig.align="center",fig.cap="Rating vs Piece Colour of winner"}
White_player=c(white_for_more,white_for_less)
Black_player=c(black_for_more,black_for_less)
Status_of_Match=c('1.More Rated Player is winner','2.Less Rated Player is winner')
datafin <- data.frame(White_player,Black_player,Status_of_Match)
datafin %>% pivot_longer(cols = !Status_of_Match, names_to = "Winner_", values_to = "Percentage") %>% 
 ggplot(aes(fill = Winner_, x = Status_of_Match, y = Percentage )) +
 geom_bar(position = "stack", stat = "identity")+
  scale_fill_manual(values = c( "slategray4","bisque2")
  )
```
The data for figure 6 shows that 61.58\% winners were the player who has more rating and among them (32.56\% ) were players with white piece.

---

## Results 

  * From dataset, including 20058 matches, it is observed that the users play rated matches more than unrated.
  * The average number of turns per match in the data set is 60.466.
  * More number of matches are won by player with white coloured piece.
  * A large number of matches ended due to resignation of any player followed by , winning by checkmate and draw.
  * _taranga_ is the user who has played the most number of matches with winning percent 46.75\%.
  * _King\'s Pawn_ is the most used opening move followed by _Queen's Pawn_, _Zukertot Opening_, _English Opening_ and _Van 't Kruijs Opening_.
  _King's Pawn_ as an opening contributes approximately in 60-63\%, _Queen's Pawn_ contributes 22\% and  _English Opening_ contributes 4\%  of the match winner\'s matches irrespective of winner\'s colour.
  This shows that whether or not the match is won by player with white coloured piece the most preferred opening is _King's Pawn_.
  * The maximum rated player has a rating of 2700, least rated player has rating 784 and mean of the ratings is 1597.
  * It is observed that high rated players won more irrespective of their colour.
  
---
  
## Conclusion

The pie chart of the _rated vs unrated matches_ gave the proportion of rated matches which contributed to change in ranking of the players.The pie chart and stacked bar graph used to show the most common opening moves also showed the relation of the opening move with the winner of the match.Both the graphs showed the preference of King's Pawn as the opening irrespective of colour of the winning piece.The pie chart of _winners per colour_ gave the percentages of matches won, showing that the player with white coloured piece won the most and the _Rating vs Piece Colour of winner_ stack chart furthermore showed that irrespective of colour of the player's piece, player with more rating won more matches.

---


\begin{center}

*      *      * 
     
\end{center}

